FLM for growth as function of SPEI

### 2021-01-19 analysis/paper/vital rate models.Rmd
switched gam() to bam() for speed.  Added year as a random effect.  Attempt at plotting covariates (but not correct yet).

- Switch measure of plant size to shoots * height
- ~~Change lag to 36 months (3 years)~~
- Produce ridge plot

@@

```{r}
demog_full <-
  demog_full %>% 
  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor))
# View(demog_full)
```

@@

@@

```{r}
g_post_cf <-
  gam(log_size_next ~ 
        s(log_size) + 
        s(plot, bs = "re") + #random effect of plot
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #unimodal-ish response to drought, but one knot per month lag.
          xt = list(bs = "cr")),
      # family = gaussian(link = "identity"),
      family = scat(link = "identity"), #like leptokurtic gaussian.
      data = demog_post_cf,
      # gamma = 1.2, #I maybe need to optimize this with cross-validation?
      method = "REML")
```

@@

```{r}
AIC(g_post_cf)

appraise(g_post_cf)
```

@@

```{r}
plot(g_post_cf, select = 1, main = "select = FALSE")
```

Hmm. Make sure this doesn't predict negative sizes.

```{r}
summary(g_post_cf)
```

Remember, R-sq is probably so high because size this year is a very good predictor of size next year.

```{r}
g_post_1ha <-
  gam(log_size_next ~ 
        s(log_size) +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #unimodal response to drought, but one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "identity"), #like leptokurtic gaussian.
      data = demog_post_1ha,
      # gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```

@@

```{r}
summary(g_post_1ha)
```

```{r}
plot(g_post_1ha, select = 1)
```

@@

```{r}
# use mean size plant across both habitats so plots are comparable
ref_data <-
  demog_post %>%
  dplyr::summarize(log_size = log(mean(size, na.rm = TRUE)),  #log of mean size, not mean of log_size
                   plot = "newlevel")
# exp(1.032814) # 2.8 shoots
mean_size <- round(exp(ref_data$log_size), 1)
```

```{r}
pred_g_post_cf  <- cb_margeff(g_post_cf, Q = spei_history, L = L, ref_data = ref_data)
pred_g_post_1ha <- cb_margeff(g_post_1ha, Q = spei_history, L = L, ref_data = ref_data)
```

@@

**Raster plots:**

```{r}
p_growth_cf <-
  pred_g_post_cf %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous(
    "lag (months)",
    breaks = 0:36, 
    expand = c(0,0),
    sec.axis = dup_axis(
      name = "Month",
      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
  ) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("#shoots (t+1)", option = "A", limits = exp(shoot_lim)) +
  labs(title = "Continuous Forest",
       caption = glue("Predicted values for average sized plant ({mean_size} shoots)")) +
  theme_bw()
p_growth_cf
```

```{r}
p_growth_1ha <-
  pred_g_post_1ha %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous(
    "lag (months)",
    breaks = 0:36, 
    expand = c(0,0),
    sec.axis = dup_axis(
      name = "Month",
      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
  ) +  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("#shoots (t+1)", option = "A", limits = exp(shoot_lim)) +
  labs(title = "1-ha Fragments",
       caption = glue("Predicted values for average sized plant ({mean_size} shoots)")) +
  theme_bw()
p_growth_1ha
```

@@

**Slice through individual months**

@@

```{r}
s_post_cf <-
  gam(surv ~ 
        s(log_size) +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_cf,
      # gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```
```{r}
# gam.check(s_post_cf) #not sure what to expect from any of these plots
appraise(s_post_cf, method = "simulate")
```
```{r}
summary(s_post_cf)
```
wow, not significant!

```{r}
s_post_1ha <-
  gam(surv ~ 
        s(log_size) +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_1ha,
      # gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```
```{r}
summary(s_post_1ha)
```
not significant!  I should try select=TRUE here.

@@

```{r}
pred_s_post_cf  <- cb_margeff(s_post_cf,  Q = spei_history, L = L, ref_data = ref_data)
pred_s_post_1ha <- cb_margeff(s_post_1ha, Q = spei_history, L = L, ref_data = ref_data)
```

@@

```{r}
p_surv_cf <-
  pred_s_post_cf %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous(
    "lag (months)",
    breaks = 0:36, 
    expand = c(0,0),
    sec.axis = dup_axis(
      name = "Month",
      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
  ) + 
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
  labs(title = "Continuous Forest",
       caption = glue("Predicted values for average sized plant ({mean_size} shoots)")) +
  theme_bw()
p_surv_cf
```

```{r}
p_surv_1ha <-
  pred_s_post_1ha %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous(
    "lag (months)",
    breaks = 0:36, 
    expand = c(0,0),
    sec.axis = dup_axis(
      name = "Month",
      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
  ) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
  labs(title = "1-ha Fragments",
       caption = glue("Predicted values for average sized plant ({mean_size} shoots)")) +
  theme_bw()
p_surv_1ha
```

@@

```{r}
f_post_cf <-
  gam(flwr ~ 
        s(log_size) +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_cf,
      # gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```

```{r}
appraise(f_post_cf)
```

@@

```{r}
plot(f_post_cf, select = 1, trans = plogis)
```

Plants with exp(3) = 20 shoots have close to a 100% chance of flowering.

```{r}
f_post_1ha <-
  gam(flwr ~ 
        s(log_size) +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_1ha,
      # gamma = 1.2, #increases smoothness when above 1
      method = "REML")
```

@@

```{r}
plot(f_post_1ha, select = 1, trans = plogis)
```

@@

```{r}
pred_f_post_cf  <- cb_margeff(f_post_cf,  Q = spei_history, L = L, ref_data = ref_data)
pred_f_post_1ha <- cb_margeff(f_post_1ha, Q = spei_history, L = L, ref_data = ref_data)
```

>>>>

- ~~Switch measure of plant size to shoots * height~~
- ~~Change lag to 36 months (3 years)~~
- Produce ridge plot
- Heatmap for growth should be ∆size
- try select = TRUE (after you understand it better? blog about it first)

@@

```{r}
demog_full <-
  demog_full %>% 
  mutate(across(c(ranch, bdffp_reserve_no, plot, habitat, ha_id_number), as.factor)) %>% 
  mutate(year_fac = as.factor(year))
# View(demog_full)
```

@@

# Speed up models with bam()

I'm going to start fitting these with `bam()` to speed things up.  This sets up the paralellization.  CAUTION: you may need to change these settings for your machine.

```{r}
cl <- parallel::makeForkCluster()
```

@@

```{r}
g_post_cf <-
  bam(log_size_next ~ 
        s(log_size) + 
        s(year_fac, bs = "re") + 
        s(plot, bs = "re") + #random effect of plot
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #unimodal-ish response to drought, but one knot per month lag.
          xt = list(bs = "cr")),
      # family = gaussian(link = "identity"),
      family = scat(link = "identity"), #like leptokurtic gaussian.
      data = demog_post_cf,
      method = "REML",
      cluster = cl)
```

@@

```{r}
AIC(g_post_cf)
appraise(g_post_cf)
```

@@

```{r}
summary(g_post_cf)
```
Remember, R-sq is probably so high because size this year is a very good predictor of size next year.

```{r}
draw(g_post_cf)
```

```{r}
g_post_1ha <-
  bam(log_size_next ~ 
        s(log_size) +
        s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #unimodal response to drought, but one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "identity"), #like leptokurtic gaussian.
      data = demog_post_1ha,
      method = "REML",
      cluster = cl)
```

@@

```{r}
summary(g_post_1ha)
```
```{r}
draw(g_post_1ha)
```

## Covariate plots

Effect of size with all else held average.

```{r}
range(demog_post_cf$log_size)

newdata <- 
  tibble(log_size = seq(0, 7.8, length.out = 100),
         plot = "newplot",
         year_fac = "newyear", 
         spei_history = mean(demog_post_cf$spei_history),
         L = demog_post_cf$L[1:100, ])

out <- augment(g_post_cf, newdata = newdata, se_fit = TRUE)

covar_growth_cf <-
  ggplot(out, aes(x = exp(log_size), y = exp(.fitted))) + 
  geom_point(data = demog_post_cf, aes(x = size, y = size_next), alpha = 0.1) +
  geom_ribbon(aes(ymin = exp(.fitted - .se.fit * 1.96), ymax = exp(.fitted + .se.fit * 1.96)), alpha = 0.3) +
  geom_line(color = "blue") +
  # geom_rug(data = demog_post_cf, aes(x = size, y = size_next), alpha = 0.25) +
  theme_linedraw() +
  labs(x = "size in year t (shoots * height)", y = "size in year t+1 (shoots * height)")
covar_growth_cf
```
```{r}
range(demog_post_1ha$log_size)

newdata <- 
  tibble(log_size = seq(0, 7.6, length.out = 100),
         plot = "newplot",
         year_fac = "newyear",
         spei_history = mean(demog_post_1ha$spei_history),
         L = demog_post_cf$L[1:100, ])
out <- augment(g_post_1ha, newdata = newdata, se_fit = TRUE)

covar_growth_1ha <-
  ggplot(out, aes(x = exp(log_size), y = exp(.fitted))) + 
  geom_point(data = demog_post_1ha, aes(x = size, y = size_next), alpha = 0.1) +
  geom_ribbon(aes(ymin = exp(.fitted - .se.fit * 1.96), ymax = exp(.fitted + .se.fit * 1.96)), alpha = 0.3) +
  geom_line(color = "blue") +
  # geom_rug(data = demog_post_1ha, aes(x = size, y = size_next), alpha = 0.25) +
  theme_linedraw() +
  labs(x = "size in year t (shoots * height)", y = "size in year t+1 (shoots * height)")
covar_growth_1ha
```

@@

```{r}
# use mean size plant across both habitats so plots are comparable
ref_data <-
  demog_post %>%
  dplyr::summarize(log_size = log(mean(size, na.rm = TRUE)),  #log of mean size, not mean of log_size
                   year_fac = "newyear",
                   plot = "newlevel")
# exp(1.032814) # 2.8 shoots
mean_size <- round(exp(ref_data$log_size), 1)
```

```{r}
pred_g_post_cf  <- cb_margeff(g_post_cf, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
pred_g_post_1ha <- cb_margeff(g_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
```

@@

TODO: make color be ∆size (i.e. fitted - mean size from ref_data)

```{r}
p_growth_cf <-
  pred_g_post_cf %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous(
    "lag (months)",
    breaks = 0:36, 
    expand = c(0,0),
    sec.axis = dup_axis(
      name = "Month",
      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
  ) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("size(t+1)", option = "A", limits = exp(shoot_lim)) +
  labs(title = "Continuous Forest") +
  theme_bw()
p_growth_cf
```

```{r}
p_growth_1ha <-
  pred_g_post_1ha %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = exp(fitted), fill = exp(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous(
    "lag (months)",
    breaks = 0:36, 
    expand = c(0,0),
    sec.axis = dup_axis(
      name = "Month",
      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
  ) +  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("#shoots (t+1)", option = "A", limits = exp(shoot_lim)) +
  labs(title = "1-ha Fragments") +
  theme_bw()
p_growth_1ha
```

@@

**Slice through individual months**
TODO: Change to ridge plots?

@@

```{r}
s_post_cf <-
  bam(surv ~ 
        s(log_size) +
        s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_cf,
      method = "REML",
      cluster = cl)
```
```{r}
# gam.check(s_post_cf) #not sure what to expect from any of these plots
appraise(s_post_cf)
```
```{r}
summary(s_post_cf)
```

```{r}
draw(s_post_cf)
```

```{r}
s_post_1ha <-
  bam(surv ~ 
        s(log_size) +
        s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_1ha,
      method = "REML",
      cluster = cl)
```
```{r}
summary(s_post_1ha)
```

not significant!  I should try select=TRUE here.

```{r}
draw(s_post_1ha)
```

## Covariate plots

Effect of size with all else held average.

```{r}
range(demog_post_cf$log_size)
newdata <- 
  tibble(log_size = seq(0, 7.8, length.out = 100),
         plot = "newplot",
         year_fac = "newyear",
         spei_history = mean(demog_post_cf$spei_history),
         L = demog_post_cf$L[1:100, ])
out <- augment(s_post_cf, newdata = newdata, se_fit = TRUE)

covar_s_cf <-
  ggplot(out, aes(x = log_size, y = plogis(.fitted))) + 
  # geom_point(data = demog_post_cf, aes(x = size, y = size_next), alpha = 0.1) +
  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit * 1.96), ymax = plogis(.fitted + .se.fit * 1.96)), alpha = 0.3) +
  geom_line(color = "blue") +
  # geom_rug(data = demog_post_cf, aes(x = size, y = size_next), alpha = 0.25) +
  theme_linedraw()
covar_s_cf
```

@@

```{r}
pred_s_post_cf  <- cb_margeff(s_post_cf,  Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
pred_s_post_1ha <- cb_margeff(s_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
```

@@

```{r}
p_surv_cf <-
  pred_s_post_cf %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous(
    "lag (months)",
    breaks = 0:36, 
    expand = c(0,0),
    sec.axis = dup_axis(
      name = "Month",
      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
  ) + 
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
  labs(title = "Continuous Forest") +
  theme_bw()
p_surv_cf
```

```{r}
p_surv_1ha <-
  pred_s_post_1ha %>%
  filter(min_dist < 0.1) %>%
  ggplot(aes(x = x, y = lag, z = plogis(fitted), fill = plogis(fitted))) +
  geom_raster(interpolate = TRUE) +
  scale_y_continuous(
    "lag (months)",
    breaks = 0:36, 
    expand = c(0,0),
    sec.axis = dup_axis(
      name = "Month",
      labels = month(ymd("2020-2-01") - months(0:36), label = TRUE))
  ) +
  scale_x_continuous("SPEI_3", expand = c(0,0)) +
  scale_fill_viridis_c("P(survival)", option = "A", limits = plogis(surv_lim)) +
  labs(title = "1-ha Fragments") +
  theme_bw()
p_surv_1ha
```

@@

```{r}
f_post_cf <-
  bam(flwr ~ 
        s(log_size) +
        s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_cf,
      method = "REML",
      cluster = cl)
```

```{r}
appraise(f_post_cf)
```
these residuals are troubling

@@

```{r}
draw(f_post_cf)
```

```{r}
f_post_1ha <-
  bam(flwr ~ 
        s(log_size) +
        s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_1ha,
      method = "REML",
      cluster = cl)
```

@@

```{r}
draw(f_post_1ha)
```

@@

```{r}
pred_f_post_cf  <- cb_margeff(f_post_cf,  Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
pred_f_post_1ha <- cb_margeff(f_post_1ha, Q = spei_history, L = L, ref_data = ref_data, cluster = cl)
```
